---
title: "peptide identification using protViz"
author: "Christian Panse / Jonas Grossmann"
date: "2016/01/24"
output: 
  - slidy_presentation
  - rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{peakplot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# FASTA - get current SwissProt

```{bash echo=FALSE}
export PATH=$PATH:~/bin/
```

```{bash}
test -f swissprot \
|| curl ftp://ftp.uniprot.org/pub/databases/uniprot/\
current_release/knowledgebase/complete/uniprot_sprot.fasta.gz \
| gunzip -c \
> swissprot
```


```{bash}
head swissprot
```

# Top ten organisms having the largest number of entries in swissprot
```{bash}
grep "^>" < swissprot \
| awk '$1~/_/{print $1}' \
| awk -F"_" '{h[$NF]++}END{for (i in h){print i" "h[i]}}' \
| sort -k2 -g -r -S128M \
| head -n 10
```

#tryptic digest rule

cut after R or K if the C-term. residue is not P.

```{bash}

```

# MS detectable peptides

tryptic digest rule: cut after R or K if the C-term. residue is not P.

```{bash}
grep "^>sp|P21781|FGF7_HUMAN" -A 5 < swissprot \
| ~/bin/fcat \
| ~/bin/digest \
| ~/bin/pim  \
| awk -F";" '600<$2 && $2<6000{print }' 

```


# CRAN package  protViz

https://CRAN.R-project.org/package=protViz

* Most frequently used R-code in the last 10 yrs

* Tailored to our site (e.g.  proprietary software exports)

* Slim, minimal dependencies for other R-packages

* Used for prototyping, teaching, analysis, or just fun

* install R

http://stat.ethz.ch/CRAN/

```{r eval=FALSE}
install.packages('protViz')
```

#  install ISCB2017

* contains data and code for hands-on-session

```{r eval=FALSE}
# urlget
install.packages('ISCB2017')
```

```{r echo=TRUE}
library(ISCB2017)
```

```

# Amino Acids (AA)

```{r}
AA
```


# Determine peptide weight 

```PANSER``` and ```GROSSMANNR```

* n_term weight?

* c_term weight?

# Swissprot SP (TASK)

- count peptides of ```swissprot_tryptic_peptides10K```

- compute peptide weight using the function ```parentIonMass```

- plot the peptide weight distribution using ```hist```



# Compute mass windows

```{r}
peptideMassTolerancePPM <- c(1, 5, 10, 20, 50)
```

```{r}
eps <- 1000 * peptideMassTolerancePPM * 1E-06
```

# Determine the number of peptide candidates

```{r fig.retina=3}
library(ISCB2017)
pim <- parentIonMass(swissprot_tryptic_peptides10K)
hist(pim)

massTolDa <- c(1E-4, 1E-3, 1E-2, 1E-1, 1, 3)
rv.Da <- lapply(massTolDa, function(eps){
  q <- runif(1000, 300,6000)
  lower <- findNN_(q - eps, pim)
  upper <- findNN_(q + eps, pim)

  data.frame(massTol = eps, n = 1 + upper-lower)
  
})


massTolPPM <- c(1, 5, 10, 20, 50)
rv.PPM <- lapply(massTolPPM, function(peptideMassTolerancePPM){
  query.mass <- runif(1000, 300,6000)
  eps <- query.mass * peptideMassTolerancePPM * 1E-06
  

  lower <- findNN_(query.mass - eps, pim)
  upper <- findNN_(query.mass + eps, pim)
  
  data.frame(massTolPPM = peptideMassTolerancePPM, n = 1 +upper - lower)
})

op <- par(mfrow=c(1,2))
boxplot(n ~ massTol,log='y', 
  data=do.call('rbind', rv.Da), 
  xlab='mass tolerance [in Da]',
  ylim=c(1,2E+5),
  ylab='number of tryptic SP peptides mass range [300,6000]')

boxplot(n ~ massTolPPM, log='y', 
  data=do.call('rbind', rv.PPM),
  xlab='mass tolerance [in ppm]',
  ylim=c(1,2E+5),
  ylab='number of tryptic SP pepitdes mass range [300,6000]')
```


# define the ions

```{r}
byIon <- function(b, y){
    Hydrogen <- 1.007825
    Oxygen <- 15.994915
    Nitrogen <- 14.003074

    yo <- y - Oxygen - Hydrogen - Hydrogen
    
    c <- b + (Nitrogen + (3 * Hydrogen))
    z <- y - (Nitrogen + (3 * Hydrogen))
  
    c2_ <- (c + Hydrogen) * 0.5
    z2_ <- (z + Hydrogen) * 0.5
  
    b2_ <- (b + Hydrogen) * 0.5
    y2_ <- (y + Hydrogen) * 0.5

    return(cbind(b, y, b2_, y2_))#, c, z, c2_, z2_, yo))
}
```

# Compute fragmention table of 'EGVNDNEEGFFSAR'

```{r}
fragmentIon('TAFDEAIAELDTLNEESYK', FUN=byIon)
```

# Compute a peptide spectrum match (psm)

```{r}
# Label tandem MS for protViz
data(msms)
plot(msms[[1]]$mZ, msms[[1]]$intensity, type='h')

# Compute a peptide spectrum match 
psm("TAFDEAIAELDTLNEESYK", msms[[1]])
```





## TAFDEAIAELDTLNEESYK

```{r fig.retina=3}
rv <- peakplot("TAFDEAIAELDTLNEESYK", msms[[1]])
```

## TAFDEAIAELDTLSEESYK

```{r fig.retina=3}
rv <- peakplot("TAFDEAIAELDTLSEESYK", msms[[2]])


#peakplot(peptide=F161375[[160]]$peptideSequence, 
#         spec=F161375[[160]],
#         FUN=function(b,y){
#           return (list(y=y))
#         })
```


# example

```{r}



peptide_identification



idx <- which(sapply(F244088f, function(x) {
length(x$mZ) > 0
}))

S <- do.call('rbind', lapply(F244088f[idx], function(x) {
  peptide_identification(
    x,
    swissprot_tryptic_peptides10K,
    pim,
    peptideMassTolerancePPM = 14,
    FUN = byIon
  )
}))


```

