---
title: "FDR - false discovery rate estimation from decoys"
author: "Witold Wolski"
date: "January 21, 2017"
output: 
  - slidy_presentation
  - rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LabelFreeQuantification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

---


```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
library(plyr)
library(ISCB2017)

```

# Overview MS

- proteotypic peptides 
- FDR as function of score
- finding score given FDR
- various FDR levels
- protein inference problem

# Overview R

- data frames (factor and character) stringAsFactors = FALSE
- tabulating data `table`
- subsetting data frames using `subset` and `[]`
- aggregating data using plyr (`ddply`)
- histograms  `hist`
- comparing lists using `intersect` and `setdiff`
- the `with` function

# Peptide annotation

```{r}
data("peptide_proteinAnnotation")
head(peptide_proteinAnnotation)
prots <- peptide_proteinAnnotation

```

# Number of Proteins per Protein


```{r fig.cap="number of PSMs per protein"}

#prots <- read.table(file="../data/peptide_proteinAnnotation.txt")
tmp <- table(prots$proteinID)
plot(table(prots$proteinID), ylab="# of PSM/protein", xlab="protein", axes=F)
axis(1)
axis(2)
names(tmp)[1:10]

```

# Number of proteins matched by one peptide

```{r fig.cap="nr. of proteins matched by one peptide"}
tmp <- table(prots$peptideSequence)
plot(sort(tmp), axes=F, ylab="# of protein per peptide", xlab="unique peptide")
axis(1)
axis(2)

```

# Length of peptide vs number of proteins matched

```{r fig.cap="length of peptide vs # proteins matched"}

x<-nchar(names(tmp))
plot(x,tmp, xlab="length of peptide sequence", ylab="nr of proteins assigned",axes=F)
axis(1)
axis(2)

```


```{r}
as.character(prots[ prots$peptideSequence == names(tmp[x>30 & tmp >50]),"proteinID"])

```

# Looking at the peptide spectrum matches for the yeast dataset

Load data.

```{r}
# specMeta <- read.table(file = "../data/psmMatchesAnnotated.txt",stringsAsFactors = F)
data("psmMatchesAnnotated")
lapply(psmMatchesAnnotated,class)
for(i in 1:ncol(psmMatchesAnnotated)){
  if(class(psmMatchesAnnotated[[i]])=="factor"){
    psmMatchesAnnotated[[i]]<-as.character(psmMatchesAnnotated[[i]])
  }
}
lapply(psmMatchesAnnotated,class)

specMeta <- psmMatchesAnnotated
table(specMeta$fileName)
```


```{r}
tmp <- specMeta$fileName%in%unique(specMeta$fileName)[1]
Glucdata <- specMeta[tmp,]
tmp <- specMeta$fileName%in%unique(specMeta$fileName)[2]
Ethanodata <- specMeta[tmp,]

```

# Compute function of FDR given score

```{r}

computeFDRFunction <- function(score, proteinID, decoy = "REV_" ){
  idx <- order(score, decreasing = TRUE)
  score <- score[idx]
  decoy_hit <- grepl(decoy, proteinID[idx])
  
  FP <- cumsum(decoy_hit)
  TP <- 1:length(idx) - FP
  
  FDR1 <- (2 * FP) / (TP + FP)
  FDR2 <- FP / TP
  return(list(decoy_hit = decoy_hit, score = score , FDR1 = FDR1 ,FDR2 = FDR2))
}

```

# Helper function to make histogram of the decoy (FP) and non decoy (TP + FP) scores.

```{r}

histWithFDR <- function(data){
  tx <- with(data,hist(score2,plot = FALSE, breaks=100))
  t1<-with(data,hist(score2[!grepl("REV_",proteinID)],breaks=tx$breaks,main="",xlab="score"))
  with(data,hist(score2[grepl("REV_",proteinID)],add=T,breaks = t1$breaks, col=2))
  cff <- with(data,computeFDRFunction(score2 , proteinID))
  par(new=T)
  with(cff,plot(score,FDR1*100, type="l",col=4,lwd=2,xlab=NA,ylab=NA,axes=FALSE))
  axis(side = 4)
  mtext(side = 4, line = 3, 'Number genes selected')
  return(cff)
}

```

# PSM level FDR vs Score

```{r}
cff <- histWithFDR(Glucdata)
with(cff,summary(score[decoy_hit]))
with(cff,summary(score[!decoy_hit]))

```

# FDR is a function of the score $FDR = f(score)$.

- Controlling FDR means finding the inverse. 
- Find score given an FDR $score = i(FDR)$ and remove all PSM's with lower score. 

```{r}
fdr = 0.01
scthresh <-min(cff$score[cff$FDR1 < fdr])
scthresh
```


# Filter dataset given score

```{r}
Glucdata1 <- subset(Glucdata, score2 > scthresh)
```


## What is the FDR before and after filtering?

```{r getFDR}
getFDR <- function(data){
  (2 * sum(grepl("REV_",data$proteinID ))) / nrow(data)
}
```


```{r}
getFDR(Glucdata)*100
getFDR(Glucdata1)*100
cff <- histWithFDR(Glucdata1)
with(cff,summary(score[decoy_hit]))
with(cff,summary(score[!decoy_hit]))

```


# Precursor level FDR

```{r}
GlucdataPrec <- ddply(Glucdata, .(peptideModSeq, precursorCharge),function(x){x[which.max(x$score2),]})
getFDR(GlucdataPrec) * 100

```



```{r plotPrecursorLevel, fig.cap="Precursor level score distribution"}

cff <- histWithFDR(GlucdataPrec)
with(cff,summary(score[decoy_hit]))
with(cff,summary(score[!decoy_hit]))

```

# Exercise : Control the Precursor level FDR on 1% level.



# Modsequence level FDR


```{r}
GlucdataModPep <- ddply(Glucdata, .(peptideModSeq),function(x){x[which.max(x$score2),]})
getFDR(GlucdataModPep) * 100

```



```{r plotModPeptideLevel, fig.cap="Peptide level score distribution"}

cff <- histWithFDR(GlucdataModPep)
with(cff,summary(score[decoy_hit]))
with(cff,summary(score[!decoy_hit]))

```


# Peptide Level FDR


```{r}
GlucdataPep <- ddply(Glucdata, .(peptideSeq),function(x){x[which.max(x$score2),]})
getFDR(GlucdataPep) * 100

```


```{r plotPeptideLevel, fig.cap="Peptide level score distribution"}
cff <- histWithFDR(GlucdataPep)
with(cff,summary(score[decoy_hit]))
with(cff,summary(score[!decoy_hit]))

```

# Protein Level FDR

```{r}
GlucdataProt <- ddply(Glucdata, .(proteinID), function(x){data.frame(x[which.max(x$score2),],nrPSM=nrow(x))  })
getFDR(GlucdataProt) * 100
nrow(GlucdataProt)
```

```{r proteinLevel, fig.cap="Protein level score distribution"}
cff <- histWithFDR(GlucdataProt)
with(cff,summary(score[decoy_hit]))
with(cff,summary(score[!decoy_hit]))

```

# Control FDR on protein level using score

```{r echo=FALSE}
fdr = 0.01
scthresh <-min(cff$score[cff$FDR1 < 0.01])
scthresh

```


```{r echo=FALSE}
GlucdataProt0.01 <- subset(GlucdataProt, score2 > scthresh)
nrow(GlucdataProt0.01)
getFDR(GlucdataProt0.01) * 100
```




# Removing Porteins identified by a single PSM

Improves FDR by removing "single hit wonders".

```{r proteinLevel2, fig.cap="Protein level score distribution"}
GlucdataProt2 <- (subset(GlucdataProt, nrPSM > 1))
nrow(GlucdataProt2)
getFDR(GlucdataProt2) * 100

cff <- histWithFDR(GlucdataProt2)
with(cff,summary(score[decoy_hit]))
with(cff,summary(score[!decoy_hit]))

```

# Compare results

Compare portein lists obtained by removing single hit wonders with those by 
controlling fdr using score.

```{r}
length(intersect(GlucdataProt2$proteinID, GlucdataProt0.01$proteinID))
length(setdiff(GlucdataProt2$proteinID, GlucdataProt0.01$proteinID))
length(setdiff(GlucdataProt0.01$proteinID,GlucdataProt2$proteinID ))

```

# Questions?

